Distance Sampling in \proglang{R}

### 2016-06-08 paper.Rnw
now in knitr format

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{David L Miller, Eric Rexstad, Len Thomas, Laura Marshall, Jeffrey L Laake}
\Plaintitle{Distance Sampling in R} %% without formatting
%\Shorttitle{\pkg{foo}: A Capitalized Title} %% a short title (if necessary)

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  David L Miller\\
  Integrated Statistics and University of St Andrews\\
  16 Sumner Street, Woods Hole, MA 02543, USA and\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{dave@ninepointeightone.net}\\
  URL: \url{http://converged.yt} \\ \\
  Eric Rexstad\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{Eric.Rexstad@st-andrews.ac.uk} \\ \\
  Len Thomas\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{len.thomas@st-andrews.ac.uk}\\
  URL: \url{http://lenthomas.org} \\ \\
  Laura Marshall\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{lhm@st-andrews.ac.uk}\\ \\
  Jeffrey L Laake\\
  Marine Mammal Laboratory\\
  Alaska Fisheries Science Center\\
  National Oceanic and Atmospheric Administration\\
  7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
  E-mail: \email{Jeff.Laake@noaa.gov}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

@@

\section{Introduction}
<<label=setup, echo=FALSE>>=
options(prompt = 'R> ')
@

@@

\begin{figure}
\begin{center}
<<label=detectdists, fig=TRUE, echo=FALSE, message=FALSE, height=5.5, width=9>>=

@@

\begin{figure}
\begin{center}
<<label=points-and-lines, fig=TRUE, echo=FALSE, message=FALSE, warning=FALSE, width=7, height=6>>=
opar <- par(mfrow=c(2,2))
# line and point transect examples
# top row points/samplers in space
# bottom row histograms

@@

\begin{figure}
\begin{center}
<<label=hn-hr-par-comp, fig=TRUE, echo=FALSE, height=5, width=10>>=
par(mfrow=c(2,4), mar=c(3.5, 3, 2, 1) + 0.1)

@@

\begin{figure}
\begin{center}
<<label=adjust-mix-comp, fig=TRUE, echo=FALSE, height=3, width=10>>=
par(mfrow=c(1,4), mar=c(3.5, 3, 3, 1) + 0.1)

@@

\begin{figure}
\begin{center}
<<label=minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig=TRUE, height=4, width=9>>=
par(mfrow=c(1, 3))
plot(minke_hn, showpoints=FALSE)
title("Minke whales", cex.main=1.5)
plot(amakihi_hr_obs, pdf=TRUE, cex=0.7)
title("Amakihi\n(observer)", cex.main=1.5)
plot(amakihi_hr_obs_mas, pdf=TRUE, cex=0.7)
title("Amakihi\n(observer+minutes)", cex.main=1.5)
@
\caption{Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model. Centre and right: plots of the probability density function for the amakihi models. Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates. Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function.}
\label{fig:minkeamakihi}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
<<label=amakihiqq, fig=TRUE, echo=FALSE, message=FALSE, height=3, width=6>>=
par(mfrow=c(1,2))
gof_ds(amakihi_hn, asp=1, cex=0.3)  # shrinking dot size to obliterate 45-deg line less
gof_ds(amakihi_hr_obs_mas, asp=1, cex=0.3)
@
\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.}
\label{amakihi-qq}
\end{center}
\end{figure}

@@

<<label=summary-table, results=tex, echo=FALSE>>=
library("knitr")
kable(summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas),
      digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
      booktabs=TRUE,
      caption="Summary for the detection function models fitted to the amakihi data. ``C-vM'' stands for Cram\\'{e}r-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error. Models are sorted according to AIC.\\label{tab:amakihi}")
@

@@

<<label=minke-abundance-table, results=tex, echo=FALSE>>=
minke_table <- summary(minke_hn)$dht$individuals$N
minke_table$lcl <- minke_table$ucl <- minke_table$df <- NULL
colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", "$\\text{CV}(\\hat{N}$)")
kable(minke_table, digits=3, format="latex", booktabs=TRUE, row.names=FALSE, escape=FALSE, caption="Summary of abundance estimation for the half-normal model for the minke whale data.\\label{minke-abund}")
@

>>>>

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{David L Miller, Eric Rexstad, Len Thomas, Laura Marshall, Jeffrey L Laake}
\Plaintitle{Distance Sampling in R} %% without formatting

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  David L Miller\\
  Integrated Statistics and University of St Andrews\\
  16 Sumner Street, Woods Hole, MA 02543, USA and\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{dave@ninepointeightone.net}\\
  URL: \url{http://converged.yt} \\ \\
  Eric Rexstad\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{Eric.Rexstad@st-andrews.ac.uk} \\ \\
  Len Thomas\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{len.thomas@st-andrews.ac.uk}\\
  URL: \url{http://lenthomas.org} \\ \\
  Laura Marshall\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{lhm@st-andrews.ac.uk}\\ \\
  Jeffrey L Laake\\
  Marine Mammal Laboratory\\
  Alaska Fisheries Science Center\\
  National Oceanic and Atmospheric Administration\\
  7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
  E-mail: \email{Jeff.Laake@noaa.gov}
}
%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

@@

\section{Introduction}
<<label=setup, echo=FALSE, include=FALSE>>=
options(prompt = 'R> ')
knitr::opts_knit$set(out.format="sweave", cache=TRUE,prompt = TRUE)
render_sweave()
@

@@

\begin{figure}
\begin{center}
<<label=detectdists, echo=FALSE, message=FALSE, fig.height=5.5, fig.width=9>>=

@@

\begin{figure}
\begin{center}
<<label=points-and-lines, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=6>>=
opar <- par(mfrow=c(2,2))
# line and point transect examples
# top row points/samplers in space
# bottom row histograms

@@

\begin{figure}
\begin{center}
<<label=hn-hr-par-comp, echo=FALSE, fig.height=5, fig.width=10>>=
par(mfrow=c(2,4), mar=c(3.5, 3, 2, 1) + 0.1)

@@

\begin{figure}
\begin{center}
<<label=adjust-mix-comp, echo=FALSE, fig.height=3, fig.width=10>>=
par(mfrow=c(1,4), mar=c(3.5, 3, 3, 1) + 0.1)

@@

\begin{figure}
\begin{center}
<<label=minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.height=4, fig.width=9>>=
par(mfrow=c(1, 3))
plot(minke_hn, showpoints=FALSE)
title("Minke whales", cex.main=1.5)
plot(amakihi_hr_obs, pdf=TRUE, cex=0.7)
title("Amakihi\n(observer)", cex.main=1.5)
plot(amakihi_hr_obs_mas, pdf=TRUE, cex=0.7)
title("Amakihi\n(observer+minutes)", cex.main=1.5)
@
\caption{Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model. Centre and right: plots of the probability density function for the amakihi models. Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates. Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function.}
\label{fig:minkeamakihi}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
<<label=amakihiqq, echo=FALSE, message=FALSE, fig.height=3, fig.width=6>>=
par(mfrow=c(1,2))
gof_ds(amakihi_hn, asp=1, cex=0.3)  # shrinking dot size to obliterate 45-deg line less
gof_ds(amakihi_hr_obs_mas, asp=1, cex=0.3)
@
\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.}
\label{amakihi-qq}
\end{center}
\end{figure}

@@

<<label=summary-table, results='asis', echo=FALSE>>=
library("knitr")
kable(summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas),
      digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
      booktabs=TRUE,
      caption="Summary for the detection function models fitted to the amakihi data. ``C-vM'' stands for Cram\\'{e}r-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error. Models are sorted according to AIC.\\label{tab:amakihi}")
@

@@

<<label=minke-abundance-table, results='asis', echo=FALSE>>=
minke_table <- summary(minke_hn)$dht$individuals$N
minke_table$lcl <- minke_table$ucl <- minke_table$df <- NULL
colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", "$\\text{CV}(\\hat{N}$)")
kable(minke_table, digits=3, format="latex", booktabs=TRUE, row.names=FALSE, escape=FALSE, caption="Summary of abundance estimation for the half-normal model for the minke whale data.\\label{minke-abund}")
@
